Figures

### 2020-06-12 Analysis/Figures_in_R.Rmd
added penguin dataset

```{r}
set.seed(42)
## Creating a 5 trait space
iris_rand <- cbind(iris[seq(from = 1, to = 150, by = 5), 1:4], rnorm(30))

## Plotting the 5 traits
pdf(file = "../Writing/Figures/figure_methods_biplot.pdf", width = 5, height = 5)
plot(iris_rand, pch = 19, xaxt = "n", yaxt = "n")
dev.off()

## Making a distance matrix
iris_dist <- dist(iris_rand)

## Making it 2D
pdf(file = "../Writing/Figures/figure_methods_2dPCO.pdf", width = 5, height = 5)
plot(cmdscale(iris_dist, k = 2), pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()

## Making a PCA
pca <- princomp(iris_rand)$scores
pdf(file = "../Writing/Figures/figure_methods_PCA1.pdf", width = 5, height = 5)
plot(pca[,1:2], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()
pdf(file = "../Writing/Figures/figure_methods_PCA2.pdf", width = 5, height = 5)
plot(pca[,3:4], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()
pdf(file = "../Writing/Figures/figure_methods_PCA3.pdf", width = 5, height = 5)
plot(pca[,c(5,1)], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()
```

>>>>

```{r}
set.seed(42)
## Creating a 5 trait space
penguins_sub_rand <- cbind(penguins_sub[seq(from = 1, to = 150, by = 5), 1:4], rnorm(30))

## Plotting the 5 traits
pdf(file = "../Writing/Figures/figure_methods_biplot.pdf", width = 5, height = 5)
plot(penguins_sub_rand, pch = 19, xaxt = "n", yaxt = "n")
dev.off()

## Making a distance matrix
penguins_sub_dist <- dist(penguins_sub_rand)

## Making it 2D
pdf(file = "../Writing/Figures/figure_methods_2dPCO.pdf", width = 5, height = 5)
plot(cmdscale(penguins_sub_dist, k = 2), pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()

## Making a PCA
pca <- princomp(penguins_sub_rand)$scores
pdf(file = "../Writing/Figures/figure_methods_PCA1.pdf", width = 5, height = 5)
plot(pca[,1:2], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()
pdf(file = "../Writing/Figures/figure_methods_PCA2.pdf", width = 5, height = 5)
plot(pca[,3:4], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()
pdf(file = "../Writing/Figures/figure_methods_PCA3.pdf", width = 5, height = 5)
plot(pca[,c(5,1)], pch = 19, xaxt = "n", yaxt = "n", ylab = "", xlab = "")
dev.off()
```

## Iris example

```{r}
set.seed(42)
## Preparing the dataset
library(palmerpenguins)

## Sorting the dataset into the correct format (to look like penguins_sub: 150*4)
penguins_sub <- penguins
class(penguins_sub) <- "data.frame"
penguins_sub <- penguins_sub[,c(1, 3:6)]

## Removing NAs
nas <- apply(penguins_sub[,-1], 1, function(x) any(is.na(x)))
penguins_sub <- penguins_sub[!nas, ]

## Selecting only 50 specimens per species
selected <- numeric()
for(species in as.character(unique(penguins_sub[,1]))) {
    specimens <- which(penguins_sub[,1] == species)
    selected <- c(selected, sample(specimens, 50))
}
penguins_sub <- penguins_sub[selected, ]

## Z transform the data
# penguins_sub[,c(2:5)] <- apply(penguins_sub[, c(2:5)], 1, scale)

## Rownames
rownames(penguins_sub) <- c(paste0(unique(penguins_sub[,1])[[1]], 1:50),
                            paste0(unique(penguins_sub[,1])[[2]], 1:50),
                            paste0(unique(penguins_sub[,1])[[3]], 1:50))
penguins_sp <- penguins_sub[, 1]
penguins_sub <- penguins_sub[,-1]

## Colnames 
colnames(penguins_sub) <- c("Culmen\nlength", "Culmen\ndepth", "Flipper\nlength", "Body\nmass")

## Morphospaces:
## No transformations
write.csv(head(penguins_sub), file = "../Writing/Figures/trait.space.no_trans.csv")

## Distance matrix
penguins_sub_dist <- dist(penguins_sub)
write.csv(round(head(as.matrix(penguins_sub_dist)[1:5, 1:5]), digits = 2), file = "../Writing/Figures/trait.space.distance.csv")

## Ordinations
penguins_sub_pca <- princomp(apply(penguins_sub, 2, scale))$scores
write.csv(round(head(penguins_sub_pca), digits = 2), file = "../Writing/Figures/trait.space.pca.csv")
penguins_sub_pco <-cmdscale(penguins_sub_dist, k = 2)

## Visualisations
## biplots
col_var <- c("orange", "blue", "green")[as.factor(penguins_sp)]
pdf(file = "../Writing/Figures/figure_methods_biplots.pdf", width = 5, height = 5)
plot(penguins_sub,
    pch = 19,
    col = col_var)
dev.off()

## PCO
pdf(file = "../Writing/Figures/figure_methods_PCO.pdf", width = 5, height = 5)

vars <- round(apply(penguins_sub_pco, 2, var)/sum(apply(penguins_sub_pco, 2, var))*100, 2)
plot(penguins_sub_pco,
    pch = 19,
    col = col_var,
    xlab = paste0("PCO 1 (", vars[1], "%)"),
    ylab = paste0("PCO 21 (", vars[2], "%)"),
    xlim = range(penguins_sub_pco),
    ylim = range(penguins_sub_pco))
dev.off()

## PCA
pdf(file = "../Writing/Figures/figure_methods_PCA.pdf", width = 5, height = 5)
vars <- round(apply(penguins_sub_pca, 2, var)/sum(apply(penguins_sub_pca, 2, var))*100, 2) 
ranges <- apply(penguins_sub_pca, 2, range)

par(mfrow = c(4,4), mar = c(1,1,1,1), oma = c(2,2,2,2))
## First row
# 1
plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
text(0,0, paste0("PC 1\n(", vars[1], "%)"), cex = 1.9)
# 2
plot(penguins_sub_pca[, c(1,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(3)
# 3
plot(penguins_sub_pca[, c(1,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
# 4
plot(penguins_sub_pca[, c(1,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(3)
axis(4)

## Second row
# 1
plot(penguins_sub_pca[, c(2,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(2)
# 2
plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
text(0,0, paste0("PC 2\n(", vars[2], "%)"), cex = 1.9)
# 3
plot(penguins_sub_pca[, c(2,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
# 4
plot(penguins_sub_pca[, c(2,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(4)

## Third row
# 1
plot(penguins_sub_pca[, c(3,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(2)
# 3
plot(penguins_sub_pca[, c(3,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
# 3
plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
text(0,0, paste0("PC 3\n(", vars[3], "%)"), cex = 1.9)
# 4
plot(penguins_sub_pca[, c(3,4)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(4)

## Fourth row
# 1
plot(penguins_sub_pca[, c(4,1)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(1)
axis(2)
# 3
plot(penguins_sub_pca[, c(4,2)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
# 3
plot(penguins_sub_pca[, c(4,3)], xaxt = "n", yaxt = "n", ylab = "", xlab = "",
    pch = 19,
    col = col_var)
axis(1)
# 4
plot(NULL, xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5), xaxt = "n", yaxt = "n", xlab = "", ylab = "")
text(0,0, paste0("PC 4\n(", vars[4], "%)"), cex = 1.9)
dev.off()

## PCA with rotation
pdf(file = "../Writing/Figures/figure_methods_PCA_rotation.pdf", width = 5, height = 5)

peguins_ordination <- princomp(apply(penguins_sub, 2, scale))

xaxis_loading <- peguins_ordination$loadings[2,1]
yaxis_loading <- peguins_ordination$loadings[1,2]

plot(apply(penguins_sub, 2, scale)[,c(1,2)],
    pch = 19,
    col = col_var,
    xlab = "Scaled culmen length",
    ylab = "Scaled culmen depth")

arrows(x0 = 0, y0 = 0, x1 = 1, y1 = xaxis_loading, lwd = 2.5)
arrows(x0 = 0, y0 = 0, x1 = yaxis_loading, y1 = 1, lwd = 2.5)
text(x = 1.35, y = xaxis_loading, "PC1", cex = 2)
text(x = yaxis_loading+0.05, y = 1.15, "PC2", cex = 2)

dev.off()
